The SMT Competition 2015--2018

### 2019-08-31 2018/report/SMT-COMP.tex
Minor

Internally, many SMT solvers employ SAT solving techniques to deal
with the propositional structure of the formula, and combine these
with (semi-)decision procedures for the background theories that the
solver
supports~\cite{DBLP:conf/cav/BarrettCDHJKRT11,DBLP:conf/cade/BoutonODF09,DBLP:conf/tacas/MouraB08,Nieuwenhuis:2006:SSS:1217856.1217859}.
Historically, SMT solvers have focused on quantifier-free formulas and
decidable combinations of background theories.  Increasingly, however,
SMT solvers also support quantified formulas.  Hence there is overlap
between SMT solving and automated theorem proving for quantified
Boolean formulas (QBF), first-order logic, and even higher-order
logic~\cite{DBLP:journals/corr/abs-1712-01486}.

@@

\paragraph{Benchmarks with unknown status.}
Another paradigm shift concerns the treatment of benchmarks with
unknown status.  Because of the difficulty to judge solver responses
for such benchmarks, SMT Competitions until 2015 only used benchmarks
whose status was known (i.e., that had been solved previously by
existing SMT solvers).  In 2015, \num{30717} non-incremental benchmarks
from the SMT Library were not used for the competition because their
status was unknown.  However, this approach was criticized for
imitating existing solvers rather than truly innovating.  In~2016,
the SMT Library contained \num{29724} non-incremental benchmarks with
unknown status; solver performance on these was evaluated but reported
separately.  Since 2017, all non-incremental benchmarks are eligible
for the competition's \maintrack regardless of status, and they
directly affect the competition results.  Solver responses for such
benchmarks are assumed to be correct, unless there is a disagreement
between two or more solvers that are otherwise (i.e., on benchmarks with
known status) sound.  Benchmarks with disagreements are removed from
the competition results.  This conservative approach was chosen over a
voting system as the latter is unreliable (especially in divisions
with few solvers), could lead to a solver being punished wrongly, and
could be gamed by entering multiple copies of a solver into the
competition.  Moreover, the size and time frame of the competition
demand an automated resolution.  Disagreements are reported to the
solver developers involved, who typically perform a manual
investigation at their leisure, and in all cases have led to bugs in
solvers being identified and fixed.  In 2016, there were disagreements
on 79 benchmarks; in 2017, only one benchmark was removed; while in
2018, there were no disagreements between otherwise sound solvers.

@@

The use of StarExec ensures the \emph{replicability} of the competition. The publicly accessible space \url{https://www.starexec.org/starexec/secure/explore/spaces.jsp?id=2641} stores all of the information required to rerun the competition, e.g., all benchmarks and copies of all tools. In addition, a GitHub organization (\url{https://github.com/SMT-COMP/smt-comp}) has been set up to archive previous results and store the tools used by organizers to run the competition.

@@

Further, there was no restriction on the number of solver versions participants
may submit, but the organizers reserved the right to not accept multiple
versions of a solver in case the number of solver submissions is too large for
the computational resources. In the past four years, exercising this right was
not necessary.
%
Many of these solvers were submitted in multiple years.  Additionally,
to promote a wide comparison between tools, developers are allowed to
submit multiple versions of a solver, e.g., for different competition
tracks or with differently tuned internal components.
Table~\ref{table:participants-history} shows in more detail how many
versions of each solver were submitted in each year.

@@

It can be easy for developers to create multiple versions of their
solver, which may differ only in configuration settings or other minor
details.  We caution against over-interpreting these numbers, which to
some extent depend on how experimental development teams were in any
given year.

By number of solver versions submitted, the competitions in 2015--2018
were the four largest in the history of SMT-COMP.  On average, 26
solver versions were submitted each year since 2015.  In contrast, the
competitions from 2005--2014 only received an average of 12
submissions~\cite{CDW14}.

@@

The \emph{(weighted) score} of a solver in a (\maintrack or
\ucoretrack) division~$D$ is then defined as the weighted sum of all
raw scores
%
$$\sum_{b\in D} \alpha'_b \cdot \langle e_b \cdot |D|, n_b \cdot |D|, w_b,
  c_b\rangle$$
%
where the sum is over all benchmarks in the division, and~$|D|$ is the
number of benchmarks in~$D$.  This is computed separately for parallel
and sequential scores.

@@

The competition-wide ranking for the \maintrack was retained and
refined after~2014.  Sequential and parallel performance are reported
separately.  Let $N_i$ be the total number of benchmarks in
division~$i$ that were used in the competition (and not removed
because of disagreements), and let $\langle e_i, n_i, w_i, c_i\rangle$
be a solver's raw score for this division.
%
In 2015 and~2016, the \emph{competition-wide raw score} for a solver
was defined as
%
$$\sum_i (e_i == 0 \,?\, (n_i/N_i)^2 : -e_i) \ln N_i$$
%
where the sum is over all competitive divisions into which the solver
was entered.
%
In 2017, the (per-division) penalty for erroneous results was changed
from the number of errors to a fixed constant, since the division
weight~$\ln N_i$ already takes the number of benchmarks in the
division into account, and multiplying it by the number of errors
overemphasizes large divisions.  In 2017 and 2018, the
\emph{competition-wide raw score} for a solver was thus defined as
%
$$\sum_i (e_i == 0 \,?\, (n_i/N_i)^2 : -4) \ln N_i$$
%
The value of this constant was chosen to balance the strong interest
in correct solvers against the risk of stifling innovation.  In the
division rankings (Section~\ref{sec:scoring-division-rankings}), an
incorrect response causes a solver to be ranked below all correct
solvers for that division.  But the competition's \maintrack consists
of many independent divisions; a solver bug that affects one division
need not manifest itself in other divisions.  With the current value,
entering a (possibly buggy) solver that can solve all benchmarks into
a division has a positive expected score if the probability that a
soundness bug in the solver will be triggered is below~$20\,\%$.

@@

The SMT Competition consists of several tracks with a multitude of divisions
per track.
Not all participating solvers enter all tracks, and the majority of
participants participates only in a handful of divisions.
To give a measure of overall progress in solver performance we
use the notion of \emph{virtual best solver} (VBS) over all divisions of all
three tracks.

For the \maintrack, we determine the VBS as the best wall-clock
performance per (correctly solved) instance with the time limit of \SI{1200}{s} used
in the competitions 2017 and 2018. Figure~\ref{fig:progress-all} visualizes VBS
performance on common benchmarks from 2015--2018 as cactus plot.
Figure~\ref{fig:progress} shows VBS performance on common benchmarks over pairs
of consecutive years from 2014 to 2018.  Both plots show the wall-clock solving
time per instance over all instances, sorted by solving time.
%
Between 2014 (\num{67426} benchmarks) and~2015 (\num{154238} benchmarks),
the number of common benchmarks is \num{67070};
between 2015 and 2016 (\num{154424} benchmarks) it is \num{154238};
between 2016 and 2017 (\num{238758} benchmarks) it is \num{145236};
and between 2017 and 2018 (\num{333241} benchmarks) it is \num{226429}.
The number of solved instances on these sets of common benchmarks
increased from 2014 to 2015 by 0.1\%, from 2016 to 2017 by 3.9\%,
and from 2017 to 2018 by 0.3\%.
Interestingly, it decreased from 2015 to 2016 by 3.5\%.
This is mainly due to the fact that, in 2015, more than \num{5600} benchmarks of
division QF\_FP where solved by one of the two participating configurations of
Z3 that no one else could solve, and in~2016 Z3 did not participate in this
division.

@@

The largest numbers in Table~\ref{table:unique-solutions} are due to
divisions in which there was only one sound solver: in~2016, MathSAT
was the only solver to support QF\_FP; in~2017, Z3 was the only sound
solver in QF\_FP; and in~2018, CVC4 was the only solver to support
QF\_SLIA, as well as the only sound solver in QF\_ABVFP.  Note that
divisions with fewer than two competing solvers do not affect the
competition-wide ranking.

@@

%TODO: discuss whether this can be attributed to the hardness
%distribution of benchmarks (cf. VBS plot)

In Table~\ref{table:parallelism}, we report the quotient of CPU time
over wall-clock time (summed up over all \maintrack benchmarks) for
each solver, as well as for all solvers combined.  When there were
multiple versions of a solver (such as a sequential and a parallel
version), only the highest value is shown.  We note that this quotient
is close to~1 for most solvers: only about 4--8 \maintrack solvers
each year were participating with a version that made use of
parallelism.

@@

In many cases, an SMT application produces either mainly satisfiable or mainly
unsatisfiable queries to the SMT solver.  A typical example is bounded model checking,
where a query is only satisfiable if a property of the model does not hold.  As
a consequence, some solvers may implement specialized techniques and
optimizations for the more frequent (the expected) case.
%
Therefore, it is interesting to ask whether the winners of divisions of a track
might have changed if benchmarks in that division had been restricted to one of
these results.  Tables~\ref{tab:results:sat}-\ref{tab:results:unsat} show the
changes in division winners if only satisfiable or unsatisfiable results are
given. These tables only lists the result for a division if it has changed
with respect to Table~\ref{tab:winners-main}. A non-competing solver is listed
by itself if there was no competing solver with a positive division score under
the given constraints.
It is interesting to note that we see some new solvers winning divisions that
did not win any division in the main results. For example, veriT solves the
most satisfiable benchmarks in QF\_UFIDL in 2018 and the most unsatisfiable
benchmarks in all four years in UFLRA, and SMTInterpol performs very well on
unsatisfiable benchmarks in QF\_LIA.

>>>>

Internally, many SMT solvers employ SAT solving techniques to deal
with the propositional structure of the formula, and combine these
with (semi-)decision procedures for the background theories that the
solver
supports~\cite{DBLP:conf/cav/BarrettCDHJKRT11,DBLP:conf/cade/BoutonODF09,DBLP:conf/tacas/MouraB08,Nieuwenhuis:2006:SSS:1217856.1217859}.
Historically, SMT solvers have focused on quantifier-free formulas and
decidable combinations of background theories.  Increasingly, however,
SMT solvers also support quantified formulas.  Hence there is overlap
between SMT solving and automated theorem proving for quantified
Boolean formulas~(QBF), first-order logic, and even higher-order
logic~\cite{DBLP:journals/corr/abs-1712-01486}.

@@

\paragraph{Benchmarks with unknown status.}
Another paradigm shift concerns the treatment of benchmarks with
unknown status.  Because of the difficulty to judge solver responses
for such benchmarks, SMT Competitions until 2015 only used benchmarks
whose status was known (i.e., that had been solved previously by
existing SMT solvers).  In 2015, \num{30717} non-incremental benchmarks
from the SMT Library were not used for the competition because their
status was unknown.  However, this approach was criticized for
rewarding imitation of existing solvers rather than true innovation.  In~2016,
the SMT Library contained \num{29724} non-incremental benchmarks with
unknown status; solver performance on these was evaluated but reported
separately.  Since 2017, all non-incremental benchmarks are eligible
for the competition's \maintrack regardless of status, and they
directly affect the competition results.

Solver responses for such
benchmarks are assumed to be correct, unless there is a disagreement
between two or more solvers that are otherwise (i.e., on benchmarks with
known status) sound.  Benchmarks with disagreements are removed from
the competition results.  This conservative approach was chosen over a
voting system as the latter is unreliable (especially in divisions
with few solvers), could lead to a solver being punished wrongly, and
could be gamed by entering multiple copies of a solver into the
competition.  Moreover, the size and time frame of the competition
demand an automated resolution.  Disagreements are reported to the
solver developers involved, who typically perform a manual
investigation at their leisure, and in all cases have led to bugs in
solvers being identified and fixed.  In 2016, there were disagreements
on 79 benchmarks; in 2017, only one benchmark was removed; while in
2018, there were no disagreements between otherwise sound solvers.

@@

The use of StarExec ensures the replicability of the competition. The publicly accessible space \url{https://www.starexec.org/starexec/secure/explore/spaces.jsp?id=2641} stores all of the information required to rerun the competition, e.g., all benchmarks and copies of all tools. In addition, a GitHub organization (\url{https://github.com/SMT-COMP/smt-comp}) has been set up to archive previous results and store the tools used by organizers to run the competition.

@@

Further, to promote a wide comparison between tools, there was no
restriction on the number of solver versions participants may submit,
e.g., for different competition tracks or with differently tuned
internal components.  Organizers reserved the right to not accept
multiple versions of a solver in case the number of solver submissions
is too large given the computational resources, but in the past four
years, exercising this right was not necessary.
Table~\ref{table:participants-history} shows in more detail how many
versions of each solver were submitted in each year.

@@

It can be easy for developers to create multiple versions of their
solver, which may differ only in configuration settings or other minor
details.  We caution against over-interpreting these numbers, which to
some extent depend on how adventurous development teams were in any
given year.

By number of solver versions submitted, the competitions in 2015--2018
were the four largest in the history of SMT-COMP.  On average, 26
solver versions were submitted each year since 2015.  In contrast, the
competitions from 2005--2014 only received an average of~12
submissions~\cite{CDW14}.

@@

The \emph{(weighted) score} of a solver in a (\maintrack or
\ucoretrack) division~$i$ is then defined as the weighted sum of all
raw scores
%
$$\sum_{b\in i} \alpha'_b \cdot \langle e_b \cdot N_i, n_b \cdot N_i, w_b, c_b\rangle$$
%
where the sum is over all benchmarks in the division, and~$N_i$ is the
number of benchmarks in~$i$.  This is computed separately for parallel
and sequential scores.

@@

The competition-wide ranking for the \maintrack was retained and
refined after~2014.  Sequential and parallel performance are reported
separately.  Let $N_i$ be the total number of benchmarks in
division~$i$ that were used in the competition (and not removed
because of disagreements), and let $\langle e_i, n_i, w_i, c_i\rangle$
be a solver's raw score for this division.
%
In 2015 and~2016, the \emph{competition-wide raw score} for a solver
was defined as
%
$$\sum_i (e_i == 0 \,?\, (n_i/N_i)^2 : -e_i) \ln N_i$$
%
where the sum is over all competitive divisions into which the solver
was entered.
%
In 2017, the (per-division) penalty for erroneous results was changed
from the number of errors to a fixed constant, since the division
weight~$\ln N_i$ already takes the number of benchmarks in the
division into account, and multiplying it by the number of errors
overemphasizes large divisions.  In 2017 and 2018, the
\emph{competition-wide raw score} for a solver was thus defined as
%
$$\sum_i (e_i == 0 \,?\, (n_i/N_i)^2 : -4) \ln N_i$$
%
The value~$-4$ was chosen to balance the strong interest in correct
solvers against the risk of stifling innovation.  In the division
rankings (Section~\ref{sec:scoring-division-rankings}), an incorrect
response causes a solver to be ranked below all correct solvers for
that division.  But the competition's \maintrack consists of many
independent divisions; a solver bug that affects one division need not
manifest itself in other divisions.  With this value, entering a
(possibly buggy) solver that can solve all benchmarks into a division
has a positive expected score if the probability that a soundness bug
in the solver will be triggered is below~$20\,\%$.

@@

The SMT Competition consists of several tracks with a multitude of
divisions per track.  Not all participating solvers enter all tracks,
and the majority of solvers participates only in a handful of
divisions.  To give a measure of overall progress in solver
performance we use, for each of the three tracks, the notion of a
\emph{virtual best solver} (VBS) over all divisions.

For the \maintrack, we determine the VBS as the best wall-clock
performance per (correctly solved) instance with the time limit of \SI{1200}{s} used
in the competitions 2017 and 2018. Figure~\ref{fig:progress-all} visualizes VBS
performance on common benchmarks from 2015--2018 as cactus plot.
Figure~\ref{fig:progress} shows VBS performance on common benchmarks over pairs
of consecutive years from 2014 to 2018.  Both plots show the wall-clock solving
time per instance over all instances, sorted by solving time.
%
Between 2014 (\num{67426} benchmarks) and~2015 (\num{154238}
benchmarks), the number of common benchmarks is \num{67070}; between
2015 and 2016 (\num{154424} benchmarks) it is \num{154238}; between
2016 and 2017 (\num{238758} benchmarks) it is \num{145236}; and
between 2017 and 2018 (\num{333241} benchmarks) it is \num{226429}.
The number of solved instances on these sets of common benchmarks
increased from 2014 to 2015 by 0.1\%, from 2016 to 2017 by 3.9\%, and
from 2017 to 2018 by 0.3\%.  Interestingly, it decreased from 2015 to
2016 by 3.5\%.  This is mainly due to the fact that, in 2015, more
than \num{5600} benchmarks of division QF\_FP were solved by one of
the two participating configurations of Z3 but by no other solver, and
in~2016 Z3 did not participate in this division.

@@

The largest numbers in Table~\ref{table:unique-solutions} are due to
divisions in which there was only one sound solver: in~2016, MathSAT
was the only solver to support QF\_FP; in~2017, Z3 was the only sound
solver in QF\_FP; and in~2018, CVC4 was the only solver to support
QF\_SLIA, as well as the only sound solver in QF\_ABVFP.  Note that
divisions with fewer than two competing solvers are non-competitive
and do not affect the competition-wide ranking.

@@

In Table~\ref{table:parallelism}, we report the quotient of CPU time
over wall-clock time (summed up over all benchmarks) for each
\maintrack solver, as well as for all solvers combined.  When there
were multiple versions of a solver (such as a sequential and a
parallel version), only the highest value is shown.  We note that this
quotient is close to~1 for most solvers: only about 4--8 \maintrack
solvers each year were participating with a version that made use of
parallelism.

@@

In many cases, an SMT application produces either mainly satisfiable
or mainly unsatisfiable queries to the SMT solver.  A typical example
is bounded model checking, where a query is only satisfiable if a
property of the model does not hold.  As a consequence, some solvers
may implement specialized techniques and optimizations for the more
frequent (the expected) case.
%
Therefore, it is interesting to ask whether the winners of divisions
of a track might have changed if benchmarks in that division had been
restricted to one of these results.
Tables~\ref{tab:results:sat}--\ref{tab:results:unsat} show the changes
in division winners if only satisfiable or unsatisfiable results are
considered.  These tables only lists the result for a division if it
has changed with respect to Table~\ref{tab:winners-main}.  A
non-competing solver is listed by itself if there was no competing
solver with a positive division score under the given constraints.  It
is interesting to note that we see some new solvers winning divisions
that did not win any division in the main results.  For example, veriT
solves the most satisfiable benchmarks in QF\_UFIDL in 2018 and the
most unsatisfiable benchmarks in all four years in UFLRA, and
SMTInterpol performs very well on unsatisfiable benchmarks in QF\_LIA.
